---
title: "Math Camp Lesson 3 (Day 2)"
subtitle: "Calculus"
author: "UW--Madison Political Science"
date: "August 20, 2020"
output:
  xaringan::moon_reader:
    lib_dir: libs
    #mathjax: "https://cdnjs.cloudflare.com/ajax/libs/mathjax/2.7.5/latest.js?config=TeX-MML-AM_CHTML"
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r xaringan-themer, include = FALSE}
library("here")
source(here("R/setup-lectures.R"))
```


## Agenda

1) Second Derivatives

2) Optimization

3) Partial Derivatives

4) Integrals

---
## Second Derivatives

For some purposes, you may need to know the derivative of the derivative—how fast the rate of change is changing.

--

These are known as second derivatives, denoted $\frac{d^2[f(x)]}{dx^2}$ or $f''(x)$.

--

Consider $f(x) = 5x^3 +8x^2 +2x+4$. What are its first and second derivatives?

--


\begin{align}
f'(x) = 15x^2 + 16x + 2
\end{align}

--

\begin{align}
f''(x) = 30x + 16
\end{align}

--

Higher order (third, fourth, etc.) derivatives also exist, but are rarely relevant.

---

## First and Second Derivatives Visually

```{r, echo=FALSE, cache = TRUE, out.width = "70%"}
df <- data.frame(x = seq(-100, 100, by= 0.5))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) 5*x^3 +8*x^2 +2*x+4, colour = "black", size = 1.4) + 
stat_function(fun = function(x) 15*x^2 +16*x +2, colour = "blue", 
              linetype = "dotted",
              size = 1.4) + 
stat_function(fun = function(x) 30*x +16, colour = "red", 
              linetype = "dotted", size = 1.4) + 
  theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "gray")+
geom_vline(xintercept = 0, colour = "gray")+
xlim(-10,10) +
  ylim(-500,500) +
  geom_text(aes(-0.5, -480, label = (paste(expression("f(x) = 5x"^3+"8x"^2+"2x"+4)))),
            parse = TRUE, size = 6) +
  geom_text(aes(-6, 500, label = (paste(expression("f'(x) = 15x"^2+"16x"+2)))),
            parse = TRUE, color = "blue", size = 6) +
  geom_text(aes(-8, -315, label = (paste(expression("f''(x) = 30x"+16)))),
            parse = TRUE, color = "red", size = 6)
```

---

## Concavity and Convexity

.pull-left[
```{r, echo=FALSE, cache = TRUE, out.width= '\\textwidth', out.height= "\\textheight"}
df <- data.frame(x = seq(-2, 1, by= 0.5))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) 5*x^3 +8*x^2 +2*x+4, colour = "black",
              size = 1.4)+ theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black")+
ylim(-10,20)
```
]

.pull-right[
Concavity Theorem: If the function $f(x)$ is twice differentiable at $x=c$, then the graph of $f(x)$ is convex (concave upward) at $(x,f(x))=(c,f(c))$ if $f''(c) > 0$ and concave (concave downward) if $f''(c) < 0$.

The function to the left is $f(x) = 5x^3 +8x^2 +2x+4$ and its second derivative is $f''(x) = 30x + 16$. 

When is it convex? When is it concave? 
]

---

## Concavity and Convexity

.pull-left[
```{r, echo=FALSE, cache = TRUE, out.width= '\\textwidth', out.height= "\\textheight"}
df <- data.frame(x = seq(-2, 1, by= 0.5))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) 5*x^3 +8*x^2 +2*x+4, colour = "black",
              size = 1.4)+ stat_function(fun = function(x) 30*x +16, colour = "red", linetype = "dotted",
              size = 1.4) + theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black")+
  ylim(-10,20)
```
]

.pull-right[
Concavity Theorem: If the function $f(x)$ is twice differentiable at $x=c$, then the graph of $f(x)$ is convex (concave upward) at $(x,f(x))=(c,f(c))$ if $f''(c) > 0$ and concave (concave downward) if $f''(c) < 0$.

The function to the left is $f(x) = 5x^3 +8x^2 +2x+4$ and its second derivative is $f''(x) = 30x + 16$. 

It is convex when $x \in \left(-\frac{16}{30},\infty\right)$ and concave when $x \in \left(-\infty,-\frac{16}{30}\right)$. $x =-\frac{16}{30}$ is an *inflection point*.
]


---


## Exercises: Second Derivatives

Find the first and second derivative of the expressions below:

$f(x) = 16x^3 −3x^2 +6$

$g(x) = x-x^2$

$h(x) = 4x^{-1} + 5x^{\frac{7}{2}}$

---

## Optimization (Maximization or Minimization)

In social science and in statistics, we often want to find the maximum or minimum of a function. For example:

--

-  We have specified a function that shows the utility a president gets out of some policy in terms of their chances of reelection. Policy is the input, probability of reelection is the output. 
  + We want to see when this function is maximized, or in other words, what values of policy would give the maximum chance of reelection.
  
-  Or we have a function that determines the bias in our statistical estimates of said probability of reelection. Model parameters are inputs, bias (how much the estimate deviates from the truth) is the output. 
  + We want to minimize this function, that is, find model parameters under which the bias is smallest.

--

How do we do this? Derivatives can help!

---

## Optimization: An Example

.pull-left[
```{r, echo=FALSE, cache = TRUE}
df <- data.frame(x = seq(0,10, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) 2*x^3-33*x^2+168*x, colour = "red")+ theme_minimal()+
xlab("x") + ylab("f(x)") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black")
```
]

.pull-right[

Suppose we want to find all the maxima and minima for the function $f(x) = 2x^3 - 33x^2 + 168x$ when $x \in [0,10]$.

]
---

## Optimization: An Example

.pull-left[
```{r, echo=FALSE, cache = TRUE}
df <- data.frame(x = seq(0,10, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) 2*x^3-33*x^2+168*x, colour = "red")+ theme_minimal()+
xlab("x") + ylab("f(x)") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black")+
geom_point(x = 0, y= 0)+
geom_point(x = 10, y= 2*10^3-33*10^2+168*10)
```
]

.pull-right[

Suppose we want to find all the maxima and minima for the function $f(x) = 2x^3 - 33x^2 + 168x$ when $x \in [0,10]$.

The *absolute maximum* occurs at $x = 10$, and the *absolute minimum* occurs at $x = 0$, or the endpoints. But there also appear to be a local maximum and a local minimum in between them.

]

---

## Optimization: An Example

.pull-left[
```{r, echo=FALSE, cache = TRUE}
df <- data.frame(x = seq(0,10, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) 2*x^3-33*x^2+168*x, colour = "red")+ theme_minimal()+
xlab("x") + ylab("f(x)") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") +
geom_point(x = 0, y= 0)+
geom_point(x = 4, y= 2*4^3-33*4^2+168*4) +
geom_point(x = 7, y= 2*7^3-33*7^2+168*7)+
geom_point(x = 10, y= 2*10^3-33*10^2+168*10)+
geom_segment(x = 3, xend = 5, y =  2*4^3-33*4^2+168*4, yend= 2*4^3-33*4^2+168*4, linetype = "dotdash" ) +
geom_segment(x = 6, xend = 8, y =  2*7^3-33*7^2+168*7, yend= 2*7^3-33*7^2+168*7, linetype = "dotdash" )
```
]

.pull-right[

Suppose we want to find all the maxima and minima for the function $f(x) = 2x^3 - 33x^2 + 168x$ when $x \in [0,10]$.

The *absolute maximum* occurs at $x = 10$, and the *absolute minimum* occurs at $x = 0$, or the endpoints. But there also appear to be a local maximum and a local minimum in between them.

To determine the precise location of these local maxima and minima, note that at these points, the slope of the line is flat. This means the derivative, which captures the slope of the tangent line, is $0$.

]

---
## Optimization: An Example

Armed with this insight,  we need to find $f'(x)$ and set it equal to $0$:

--

\begin{align}
0 = f'(x) = (2x^3 - 33x^2 + 168x)' =  6x^2 −66x+168
\end{align}

--

We can simply solve this equation:

--

\begin{align}
0 &= 6x^2 - 66x+168 \\
0 &= x^2 - 11x + 28
\end{align}

--

We can solve it either by *factoring*

$0 = (x^2 - 4x) + (-7x + 28) = x(x - 4) - 7(x - 4) = (x - 4) (x - 7)$

...or by using the *quadratic formula*


$x = \frac{11 \pm \sqrt{(-11)^2-4(1)(28)}}{2(1)} = \frac{11 \pm 3}{2}$

Either way, $x = 4$ or $x = 7$

---
class: center, middle
```{r, echo=FALSE, cache = TRUE, out.width = "80%"}
df <- data.frame(x = seq(0,10, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) 2*x^3-33*x^2+168*x, colour = "red")+ theme_minimal()+
xlab("x") + ylab("f(x)") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") +
geom_point(x = 0, y= 0)+
geom_point(x = 4, y= 2*4^3-33*4^2+168*4) +
geom_point(x = 7, y= 2*7^3-33*7^2+168*7)+
geom_point(x = 10, y= 2*10^3-33*10^2+168*10)+
geom_segment(x = 3, xend = 5, y =  2*4^3-33*4^2+168*4, yend= 2*4^3-33*4^2+168*4, linetype = "dotdash" ) +
geom_segment(x = 6, xend = 8, y =  2*7^3-33*7^2+168*7, yend= 2*7^3-33*7^2+168*7, linetype = "dotdash" )+
geom_segment(x = 4, xend=4, y =0, yend = 2*4^3-33*4^2+168*4, colour = "black", linetype = "dotdash")+
geom_segment(x = 7, xend=7, y =0, yend = 2*7^3-33*7^2+168*7, colour = "black", linetype = "dotdash")+
scale_x_continuous(breaks=seq(0,10,1))
```
---
## Optimization: Some Rules

If the function is relatively simple to understand or graph, it may be possible to know whether you’re dealing with a maximum or a minimum.

--

However, it may be necessary to check by taking the second derivative and evaluating it at the point where the first derivative equals $0$, which we'll call $x^*$:

--

  - Negative $(f''(x^*) < 0)$: local maximum 
  - Positive $(f''(x^*) > 0)$: local minimum 
  - Zero $(f''(x^*) = 0)$: neither a minimum or a maximum

--

Returning to our previous example, 

\begin{align}
f'(x) &= 6x^2 - 66x+168 \\
f''(x) &= 12x-66
\end{align}

--

when we evaluate the second derivatives at the local minimum and maximum, the results are $f''(4) = -18$ and $f''(7) = 18$. 

---
## An Exercise

Find the local minimum and local maximum of the function below, and check mathematically which is the minimum and which is the maximum:

$x^3−x^2+1$

---


## Multivariate Functions & Partial Derivatives

When a function takes multiple variables as inputs, it is only possible (and sometimes useful) to take the derivative with respect to one variable at a time, treating the others as constants.

--

These are known as partial derivatives, denoted $\partial$ (pronounced "del"), or $f_x(x,y)$ if you want the derivative of a function of $x$ and $y$ with respect to $x$. 

--

Consider $f(x,y,z)=4x^2y^4 +2xz^3+8y^2z^4 +8x+7y+3z+2$:

--

\begin{align}
\frac{\partial[f(x,y,z)]}{\partial x} = f_x(x,y,z) &= 8xy^4+2z^3+8 \\
\frac{\partial[f(x,y,z)]}{\partial y} = f_y(x,y,z) &= 16x^2y^3+16yz^4+7 \\
\frac{\partial[f(x,y,z)]}{\partial z} = f_z(x,y,z) &= 6xz^2+32y^2z^3+3
\end{align}

--

Why is this useful? E.g., we are modeling election turnout as a function of age and income, and we want to know how turnout changes with respect only to changes in income.


---
## Exercise: Partial Derivatives

Find the partial derivatives of the function below with respect to each variable

$g(p,q) = 8p^2q+4pq−7pq^2 +18$

---
## Partial Higher-Order Derivatives

It is possible to combine second-order (and higher) derivatives with partial derivatives. For example:

Consider $f(x, y) = 3x^3y^2$. Let's say we want to find $\frac{\partial^2}{\partial y \partial x}f(x,y)$:

\begin{align}
\frac{\partial^2}{\partial y \partial x}(3x^3y^2) &=\frac{\partial}{\partial y}((3)3x^{3-1}y^2) \\
&=\frac{\partial}{\partial y}(9x^2y^2) \\
&=(2)9x^2y^{2-1} \\
&=18x^2y
\end{align}

--

Pay attention to the denominator to learn what operations to perform, and in what order. Here, we are taking the second derivative of the entire function, but are differentiating once with respect to $x$ and then once with respect to $y$.

If instead we were given $\frac{\partial^3}{\partial x^2 \partial y}$, we would differentiate 3 times overall, once with respect to y, and then twice with respect to x.

---

## Exercise: Partial Higher-Order Derivatives

Consider again $f(x, y) = 3x^3y^2$. Find:

- $\frac{\partial^3}{\partial x^2 \partial y}$
- $\frac{\partial^3}{\partial x \partial y^2}$

---

## Integrals

The integral is the **signed area** of the region between the curve and the x-axis.

--

Signed implies that it can be positive or negative.

--

Integral is either the total area as the function extends to infinity in either direction, or the area between two selected points.

--

One common application of integrals in social science is to calculate the probability that a variable falls in a certain range (e.g., that a conflict will leave more than a 1,000 people dead)

---

## Areas

.pull-left[
```{r, echo=FALSE, cache = TRUE}
df <- data.frame(x = seq(0, 1, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) 3, colour = "black")+ theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black")+
ylim(0,4)
```
]

.pull-right[
Let's consider the function, $y = 3$, plotted to the left. What is its area under the curve from $x=0.25$ and $x=0.75$?
]


---

## Areas

.pull-left[
```{r, echo=FALSE, cache = TRUE}
df <- data.frame(x = seq(0, 1, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) 3, colour = "black")+ theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") +
geom_segment(x = 0.25, xend = 0.25, y = 0, yend = 3, colour = "red", linetype = "dotdash") +
geom_segment(x = 0.75, xend = 0.75, y = 0, yend = 3, colour = "red", linetype = "dotdash")+
ylim(0,4)
```
]

.pull-right[
Let's consider the function, $y = 3$, plotted to the left. What is its area under the curve from $x=0.25$ and $x=0.75$?

Given that this is a rectangle, the area between the function and the x-axis is $\text{area} = (0.75-0.25) \times 3 = 1.5$
]


---

## Areas

.pull-left[
```{r, echo=FALSE, cache = TRUE}
df <- data.frame(x = seq(0, 1, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x)  0.5*x+1, colour = "black")+ theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") 
```
]

.pull-right[
Let's consider a less simple function, $y = \frac{1}{2}x+1$, plotted to the left.  What is its area under the curve from $x=0.25$ and $x=0.75$?
]


---

## Areas

.pull-left[
```{r, echo=FALSE, cache = TRUE}
df <- data.frame(x = seq(0, 1, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x)  0.5*x+1, colour = "black")+ theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") +
geom_segment(x = 0.25, xend = 0.25, y = 0, yend = 0.5*(.25)+1, colour = "red", linetype = "dotdash") +
geom_segment(x = 0.75, xend = 0.75, y = 0, yend = 0.5*(.75)+1, colour = "red", linetype = "dotdash") +
geom_segment(x = 0.25, xend = 0.75, y = 0.5*(.25)+1, yend = 0.5*(.25)+1, colour = "red", linetype = "dotdash")
```
]

.pull-right[
Let's consider a less simple function, $y = \frac{1}{2}x+1$, plotted to the left.  What is its area under the curve from $x=0.25$ and $x=0.75$?

We can find the area between the function and the x-axis by taking advantage of the fact we can draw a triangle and rectangle and sum their areas. $\text{area}_{tri} = 0.0625$ and $\text{area}_{rect}= 0.5625$. Thus, the total area is $0.625$.

]

---

## Area

.pull-left[
```{r, echo=FALSE, cache = TRUE}
df <- data.frame(x = seq(0, 1, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) x-x^2, colour = "black")+ theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") +ylim(0,0.35)

```
]

.pull-right[
Let's consider an even more complicated function, $y = x-x^2$, plotted to the left.  What is its area under the curve from $x=0.25$ and $x=0.75$?
]

---

## Area

.pull-left[
```{r, echo=FALSE, cache = TRUE}
df <- data.frame(x = seq(0, 1, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) x-x^2, colour = "black")+ theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") +ylim(0,0.35) +
geom_segment(x = 0.25, xend = 0.25, y = 0, yend = .25 -(.25)^2, colour = "red", linetype = "dotdash") +
geom_segment(x = 0.75, xend = 0.75, y = 0, yend = .75 -(.75)^2, colour = "red", linetype = "dotdash") +
geom_segment(x = 0.25, xend = 0.75, y = .25 -(.25)^2, yend = .75 -(.75)^2, colour = "red", linetype = "dotdash")  
```
]

.pull-right[
Let's consider an even more complicated function, $y = x-x^2$, plotted to the left.  What is its area under the curve from $x=0.25$ and $x=0.75$?

This would require a more complex trick.
]

---
## Integral as the Limit of a Sum

Imagine dividing the region into intervals and drawing a rectangle to capture the area for each interval, with height equal to the value of the function at the left (or right) of the interval, then summing the area of those rectangles.
  
--

Let's see what happens as we add rectangles
  
---
class: center,  middle

```{r, echo=FALSE, cache = TRUE, out.width = "80%"}
df <- data.frame(x = seq(0, 1, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) x-x^2, colour = "black")+ theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") +ylim(0,0.35) +
geom_segment(x = 0.25, xend = 0.25, y = 0, yend = .25 -(.25)^2, colour = "red", linetype = "dotdash") +
geom_segment(x = 0.75, xend = 0.75, y = 0, yend = .75 -(.75)^2, colour = "red", linetype = "dotdash") +
geom_segment(x = 0.25, xend = 0.75, y = .25 -(.25)^2, yend = .75 -(.75)^2, colour = "red", linetype = "dotdash")  
```
---

class: center, middle

```{r, echo=FALSE, cache = TRUE, out.width = "80%"}
df <- data.frame(x = seq(0, 1, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) x-x^2, colour = "black")+ theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") +ylim(0,0.35) +
# first box
geom_segment(x = 0.25, xend = 0.25, y = 0, yend = .25 -(.25)^2, colour = "red", linetype = "dotdash") +
geom_segment(x = 0.5, xend = 0.5, y = 0, yend = .25 -(.25)^2, colour = "red", linetype = "dotdash") +
geom_segment(x = 0.25, xend = 0.5, y = .25 -(.25)^2, yend = .25 -(.25)^2, colour = "red", linetype = "dotdash") +
# second box
geom_segment(x = 0.5, xend = 0.5, y = 0, yend = .5 -(.5)^2, colour = "red", linetype = "dotdash") +
geom_segment(x = 0.75, xend = 0.75, y = 0, yend = .5 -(.5)^2, colour = "red", linetype = "dotdash")+
geom_segment(x = 0.5, xend = 0.75, y = .5 -(.5)^2, yend = .5 -(.5)^2, colour = "red", linetype = "dotdash")  
```
---

class: center, middle

```{r, echo=FALSE, cache = TRUE, out.width = "80%"}
df <- data.frame(x = seq(0, 1, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) x-x^2, colour = "black")+ theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") +ylim(0,0.35) +
# first box
geom_segment(x = 0.25, xend = 0.25, y = 0, yend = .25 -(.25)^2, colour = "red", linetype = "dotdash") +
geom_segment(x = .375, xend = .375, y = 0, yend = .25 -(.25)^2, colour = "red", linetype = "dotdash") +
geom_segment(x = 0.25, xend = 0.375, y = .25 -(.25)^2, yend = .25 -(.25)^2, colour = "red", linetype = "dotdash") +
# second box
geom_segment(x =  0.375, xend =  0.375, y = 0, yend =   0.375 -(0.375)^2, colour = "red", linetype = "dotdash") +
geom_segment(x =0.5, xend = 0.5, y = 0, yend = 0.375 -(0.375)^2, colour = "red", linetype = "dotdash")+
geom_segment(x = 0.375, xend = 0.5, y = 0.375  -(0.375)^2, yend = 0.375  -(0.375)^2, colour = "red", linetype = "dotdash")  +
# third box
geom_segment(x =0.5, xend = 0.5, y = 0, yend = 0.5-(0.5)^2, colour = "red", linetype = "dotdash") +
geom_segment(x = 0.625, xend = 0.625, y = 0, yend = 0.625 -(0.625)^2, colour = "red", linetype = "dotdash")+
geom_segment(x = 0.5, xend = 0.625, y = 0.5  -(0.5)^2, yend = 0.5  -(0.5)^2, colour = "red", linetype = "dotdash")+
# fourth box
geom_segment(x = 0.625, xend = 0.625, y = 0, yend = 0.625 -(0.625)^2, colour = "red", linetype = "dotdash") +
geom_segment(x = 0.75, xend = 0.75, y = 0, yend = 0.625 -(0.625)^2, colour = "red", linetype = "dotdash")+
geom_segment(x = 0.625, xend = 0.75, y =0.625 -(0.625)^2, yend = 0.625 -(0.625)^2, colour = "red", linetype = "dotdash")
```
---

## Integral as the Limit of a Sum

Imagine dividing the region into intervals and drawing a rectangle to capture the area for each interval, with height equal to the value of the function at the left (or right) of the interval, then summing the area of those rectangles.
  

Approximation improves as the intervals become smaller.

---

## Integrals

As you reduce the width of rectangles to zero, the summed areas of the rectangles converges to the area under the curve—including more and more of the area inside and less and less of the area outside.

\begin{align}
\int_a^bf(x)dx = \displaystyle\lim_{h \to 0}\sum_{i=1}^{H}f(x_i)h_i
\end{align}

--

This is read as the ''integral of $f(x)$ from $x=a$ to $x=b$ with respect to $x$.''

--

Theoretically, this approach allows to find:

  - the exact value of the area between those points for any well-behaved   function
  - a general equation for the area between any two points 
  
--

However, it is mathematically complicated to solve these using this approach
---


## Antiderivatives

The antiderivative of a function $f(x)$, denoted $F(x)$, is the function whose derivative returns the original function. Formally, $F(x)$ is a function such that: 

\begin{align}
F'(x) = f(x)
\end{align}

--

Essentially, this “unwinds” the derivative operation, or applies it backwards.

---
## The Fundamental Theorem of Calculus

The fundamental theorem of calculus relates the antiderivative and the integral.

\begin{align}
\int_a^bf(x)dx = F(b)-F(a) =  F(x)\displaystyle\rvert_a^b
\end{align}

---
## Indefinite Integrals

We've been discussing *definite* integrals or ones that are bounded.

--

We can also have *indefinite* integrals or ones that do not have specific bounds, or 


\begin{align}
\int f(x)dx = F(x) + C
\end{align}

--

This expression uses exactly the same antiderivative as the definite integral but there is no subtraction and there’s an arbitrary constant $C$ added (since that’d disappear when taking the derivative).

---

## Straightforward Integrals

There is an analogous concept of *power rule* for antiderivatives (when $n \neq -1$).

--

If $f(x) = ax^n$, then its antiderivative is $$\int f(x)dx = F(x) + C = \frac{a}{n+1}x^{n+1}+C$$

--

Some other useful antiderivatives are:

--

\begin{align}
\int adx = ax + C
\end{align}

--

\begin{align}
\int \frac{1}{x}dx = \int x^{-1}dx = ln(x) + C
\end{align}

--

\begin{align}
\int e^xdx = e^x + C
\end{align}

---
## Straightforward Integrals

Moreover, when dealing with addition and subtraction, we can separate integrals:

--

\begin{align}
\int f(x) + g(x)dx = \int f(x)dx + \int g(x)dx = F(x)+G(x)+C 
\end{align}

--

We can also pull out constants before integrating:

--

\begin{align}
\int af(x)dx = a\int f(x)dx = aF(x) + C
\end{align}

---

## Straightforward Integrals: Example 1

Consider the example $f(x) = x - x^2$. The indefinite integral is

--

\begin{align}
\int  x - x^2dx &= \int xdx-\int x^2dx  \\
&= \frac{1}{1+1}x^{1+1}-\left(\frac{1}{2+1}x^{2+1}\right) + C\\
&= \frac{1}{2}x^{2}-\frac{1}{3}x^{3} + C
\end{align}

--

Now consider the area of the curve specifically between $.25$ and $.75$

--

\begin{align}
\int_.2^.75  x - x^2dx &= \frac{1}{2}x^{2}-\frac{1}{3}x^{3}\displaystyle\rvert_.25^.75  \\
&= \left(\frac{1}{2}(.75)^{2}-\frac{1}{3}(.75)^{3}\right) - \left(\frac{1}{2}(.25)^{2}-\frac{1}{3}(.25)^{3}\right) \\
& = \frac{11}{96}
\end{align}
---

## Straightforward Integrals: Example 2

Consider the example $f(x) = 9x^2+10x+4$. The indefinite integral is

--

\begin{align}
\int  9x^2+10x+4dx &= 9\int x^2dx+10\int xdx  + \int 4dx \\
&= 9\left(\frac{1}{3}\right)x^{3}+10\left(\frac{1}{2}\right)x^{2}+ 4x +C\\
&=  3x^3 +5x^2 +4x+C
\end{align}

--

Now consider the area of the curve specifically between 2 and 5:

--

\begin{align}
\int_2^5  9x^2+10x+4dx    &= 3x^3 +5x^2 +4x\displaystyle\rvert_2^5  \\
&= (3(5)^3 +5(5)^2 +4(5))-(3(2)^3 +5(2)^2 +4(2))\\
& = 468
\end{align}

---

## Exercise: Straightforward Integrals

Find the indefinite integral of the function below, and calculate the area under the curve between 0 and 1:

$\int (2x^3-3x^2+7x+4)dx$

---

## Advanced Integrals

There are techniques for computing the integrals of more complicated functions: 

- **integration by parts** ("reverse product rule")

- **integration by substitution** ("reverse chain rule")
  
These are beyond the scope of what we have time to cover here and, for the most part, beyond the scope of what you will need to do by hand in political science.


---

class: center, inverse, middle

# End Day 4

```{css, echo=FALSE}
@media print {
  .has-continuation {
    display: block;
  }
}
```