---
title: "Math Camp Lesson 3 (Day 1)"
subtitle: "Calculus"
author: "UW--Madison Political Science"
date: "August 19, 2020"
output:
  xaringan::moon_reader:
    lib_dir: libs
    #mathjax: "https://cdnjs.cloudflare.com/ajax/libs/mathjax/2.7.5/latest.js?config=TeX-MML-AM_CHTML"
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r xaringan-themer, include = FALSE}
library("here")
source(here("R/setup-lectures.R"))
```


## Overview

Calculus evaluates the behavior of functions:
- Limits
- Rate of change
- Change in the rate of change
- Area of the region they defined on

--

Concepts from calculus underlie a wide variety of mathematics, particularly in the applied math that we use in political science

--

Calculus in political science:

- Finding the fitline with the minimal distance between predicted and observed data

- Calculating the probability density in regions of continuous distributions

- Solving for the choice that maximizes a decision maker's utility


---

## Agenda

Day 1
  - Limits
  - Derivatives

Day 2
  - More Derivatives
  - Integrals
  - Applications
  
---

## Limits

The first important idea for calculus are limits. 

--

The limit of a function characterizes its behavior given a certain input, or as an input value changes.

---

## Limits: Example 1

--

.pull-left[
```{r, echo=FALSE, cache = TRUE}
df <- data.frame(x = seq(0,4, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) 3-(x-2)^2, colour = "red")+ theme_minimal()+
xlab("x") + ylab("f(x)") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black")
```
]

.pull-right[
Let's consider the simple function, $f(x) = y = 3-(x-2)^2$, plotted to the left. What is the limit of $f(x)$ as $x$ *approaches* 2?]

 
---

## Limits: Example 1

.pull-left[
```{r, echo=FALSE, cache = TRUE}
df <- data.frame(x = seq(0,4, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) 3-(x-2)^2, colour = "red")+ theme_minimal()+
xlab("x") + ylab("f(x)") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black")+
geom_segment(x = 0, xend = 2, y =3, yend = 3, linetype = "dotdash") +
geom_segment(x = 2, xend = 2, y =0, yend = 3, linetype = "dotdash")
```
]

.pull-right[
Let's consider the simple function, $f(x) = y = 3-(x-2)^2$, plotted to the left. What is the limit of $f(x)$ as x *approaches* 2?

As $x$ approaches 2,  $f(x)$ or $y$ approaches $f(2) = 3$.  

]

---

## Limits: Example 2

.pull-left[
```{r, echo=FALSE, cache = TRUE}
df <- data.frame(x = seq(0.5,50, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) 3-1/x, colour = "red")+ theme_minimal()+
xlab("x") + ylab("g(x)") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black")
```
]

.pull-right[
Let's consider a less simple function, $g(x) = y = 3-\frac{1}{x}$, plotted to the left. What is the limit of $g(x)$ as $x$ *approaches* $\infty$?]

 
---

## Limits: Example 2

.pull-left[
```{r, echo=FALSE, cache = TRUE}
df <- data.frame(x = seq(0.5,50, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) 3-1/x, colour = "red")+ theme_minimal()+
xlab("x") + ylab("g(x)") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black")+
geom_segment(x = 0, xend = 50, y =3, yend = 3, linetype = "dotdash") 
```
]

.pull-right[
Let's consider a less simple function, $g(x) = y = 3-\frac{1}{x}$, plotted to the left. What is the limit of $g(x)$ as $x$ *approaches* $\infty$?

As $x$ approaches $\infty$, $g(x)$ approaches 3. How do we know?

]

---

## Limits: Example 2

.pull-left[
```{r, echo=FALSE, cache = TRUE}
df <- data.frame(x = seq(0.5,50, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) 3-1/x, colour = "red")+ theme_minimal()+
xlab("x") + ylab("g(x)") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black")+
geom_segment(x = 0, xend = 50, y =3, yend = 3, linetype = "dotdash") 
```
]

.pull-right[
Let's consider a less simple function, $g(x) = y = 3-\frac{1}{x}$, plotted to the left. What is the limit of $g(x)$ as $x$ *approaches* $\infty$?

As $x$ approaches $\infty$, $g(x)$ approaches 3. Why?

As $x$ gets larger, $\frac{1}{x}$ gets smaller and smaller. $$\left(\frac{1}{2} > \frac{1}{20} > \frac{1}{200} \dots\right)$$

]

---
## Limits, Formally Written

Formally, limits are expressed as:

\begin{align}
\lim_{x \to c} f(x) = L
\end{align}

This expression should be read as: ''As $x$ approaches $c$, the limit of $f(x)$ is $L$.''

--

Many times, you will see this expression written as 

$\lim\limits_{x \to c^{-}} f(x) = L$ 

or $\displaystyle\lim_{x \to c^{+}} f(x) = L$ 

A negative sign $(-)$ implies "As $x$ approaches $c$ *from the left*" 

A positive sign $(+)$ implies "As $x$ approaches $c$ *from the right*"

---

## Tips for Taking Limits

--

Simplify as much as possible.

--

Separate out the limits into distinct elements. Move constants outside the limit operator.


--

Watch out for components that $\dots$
  - $\dots$ grow very large or very small
  - $\dots$ become zero

--

Are these components in the numerator or denominator of a fraction?

--

If you can, evaluate the function at the limit.

--

For functions that are well-behaved, the limit as $x$ approaches a finite point is generally the value of the function at that point (if it exists).

---


## Finding Limits: Example 1

Let's consider $\displaystyle\lim_{x \to 2} x^2-3x +1$

--

\begin{align}
\displaystyle\lim_{x \to 2} x^2-3x +1 &= \lim_{x \to 2} x^2 -3\lim_{x \to 2}x + \lim_{x \to 2} 1 \\
&= 2^2 - 3(2)+ 1 \\
&= -1
\end{align}

---
## Finding Limits: Example 2

Now, let's consider $\displaystyle\lim_{x \to \infty} \frac{4x^4+7x^2+8}{3x^4}$

--
\begin{align}
\displaystyle\lim_{x \to \infty} \frac{4x^4+7x^2+8}{3x^4} 
&= \lim_{x \to \infty}  \frac{4x^4}{3x^4} + \lim_{x \to \infty} \frac{7x^2}{3x^4}+ \lim_{x \to \infty}  \frac{8}{3x^4} \\
&= \lim_{x \to \infty}  \frac{4}{3} + \lim_{x \to \infty} \frac{7}{3x^2}+ \lim_{x \to \infty}  \frac{8}{3x^4} \\
&= \frac{4}{3} + 0 + 0 \\
&= \frac{4}{3} 
\end{align}

--

Why does $\lim_{x \to \infty} \frac{7}{3x^2} = 0$? As $x \to \infty$, $3x^2 \to \infty$, and $\frac{7}{\infty} \to 0$.

---

## Finding Limits: Example 3

--
Let's consider $\displaystyle\lim_{x \to 0} \frac{4x^4+7x^2 +8}{3x^4}$

--

\begin{align}
&= \lim_{x \to 0} \frac{4x^4}{3x^4} + \lim_{x \to 0} \frac{7x^2}{3x^4}+ \lim_{x \to 0} \frac{8}{3x^4} \\
&= \lim_{x \to 0} \frac{4}{3} + \lim_{x \to 0} \frac{7}{3x^2}+ \lim_{x \to 0} \frac{8}{3x^4} \\
&= \frac{4}{3} + \frac{7}{3} \lim_{x \to 0} \frac{1}{x^2}+ \frac{8}{3}\lim_{x \to 0} \frac{1}{x^4} \\
&= \frac{4}{3} + \frac{7}{3} \times \frac{1}{0} + \frac{8}{3} \times \frac{1}{0} \\
&= \dots
\end{align}

---

## Finding Limits: Example 3

The limit is $\infty$.

--
\begin{align}
&= \lim_{x \to 0} \frac{4x^4+7x^2 +8}{3x^4} \\
&= \frac{4}{3} + \frac{7}{3} \lim_{x \to 0} \frac{1}{x^2}+ \frac{8}{3}\lim_{x \to 0} \frac{1}{x^4} \\
&= \infty
\end{align}


As $x$ approaches $0$, the function retains some positive value in the numerator while the denominator *positively* approaches 0. This means that you are dividing by a smaller and smaller fraction, which means the entire term is getting larger and approaches $\infty$.  

---

class: center, middle

```{r, echo=FALSE, cache = TRUE, out.width = "80%"}
df <- data.frame(x = seq(-10,10, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) (4*x^4+7*x^2+8)/(3*x^4), colour = "red")+ theme_minimal()+
xlab("x") + ylab("f(x)") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black")+ ylim(-30,30)+
geom_segment(x = 0, xend = 0, y =3, yend = 3, linetype = "dotdash") 
```

---

## Exercises

Find the following limits:

$\lim_{x \to 4} x^2-6x+4$

$\lim_{x \to 4} \frac{x^2}{3x - 2}$

$\lim_{x \to \infty} \frac{3x - 4}{x + 3}$

---
## Derivatives

--
The derivative of a function is its rate of change in the output as the value of its input changes.

--

It is the instantaneous slope of the line at any given point.

--


The slope of a function is how much the output changes as a result of changes in the input. 

Using $\Delta$ to signify ‘change’, this is 
$\frac{\Delta f(x)}{\Delta x} = \frac{f(x_2)-f(x_1)}{x_2-x_1}$ or ''rise-over-run''.

---

## Slope

.pull-left[
```{r, echo=FALSE, cache = TRUE}
df <- data.frame(x = seq(-5, 5, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) 3, colour = "red")+ theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") +
ylim(0,5)
```
]

.pull-right[
Let's consider the function, $y = 3$, plotted to the left. What is its ''slope''?
]

---

## Slope

.pull-left[
```{r, echo=FALSE, cache = TRUE}
df <- data.frame(x = seq(-5, 5, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) 3, colour = "red")+ theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") +
ylim(0,5)
```
]

.pull-right[
Let's consider the function, $y = 3$, plotted to the left. What is its ''slope''?

Its slope or $\frac{\Delta f(x)}{\Delta x} = 0$ because there is no "rise".
]

---


## Slope

.pull-left[
```{r, echo=FALSE, cache = TRUE}
df <- data.frame(x = seq(-5, 5, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) 0.5*x+1, colour = "red")+ theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black")

```
]

.pull-right[
Let's consider a less simple function, $y = \frac{1}{2}x+1$, plotted to the left. What is its ''slope''?
]

---

## Slope

.pull-left[
```{r, echo=FALSE, cache = TRUE}
df <- data.frame(x = seq(-5, 5, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) 0.5*x+1, colour = "red")+ theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") 
```
]

.pull-right[
Let's consider a less simple function, $y = \frac{1}{2}x+1$, plotted to the left. What is its ''slope''?

Its slope or $\frac{\Delta f(x)}{\Delta x} = \frac{1}{2}$. 

[Recall: $y =mx +b$ from Day 1]
]

---

## Slope

.pull-left[
```{r, echo=FALSE, cache = TRUE}
df <- data.frame(x = seq(0, 1, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) x-x^2, colour = "red")+ theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") +ylim(0,0.35)

```
]

.pull-right[
Let's consider an even more complicated function, $y = x-x^2$, plotted to the left. What is its ''slope'', or $\frac{\Delta f(x)}{\Delta x}$?
]

---

## Slope

.pull-left[
```{r, echo=FALSE, cache = TRUE}
df <- data.frame(x = seq(0, 1, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) x-x^2, colour = "red")+ theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") +ylim(0,0.35)
```
]

.pull-right[
Let's consider an even more complicated function, $y = x-x^2$, plotted to the left. What is its ''slope'', or $\frac{\Delta f(x)}{\Delta x}$?

How do we even calculate this?
]


---

## Derivatives as Limits

We can approximate the slope at a certain location by picking a point nearby on the line and finding the slope of the straight line connecting these two points.

--

Let's consider a few examples.

---

class: center, middle
```{r, echo=FALSE, cache = TRUE, out.width = "80%"}
df <- data.frame(x = seq(0, 1, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) x-x^2, colour = "red")+
geom_point(x=0.3, y = .21)+
theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") +ylim(0,0.35)
```
---

class: center, middle
```{r, echo=FALSE, cache = TRUE, out.width = "80%"}
df <- data.frame(x = seq(0, 1, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) x-x^2, colour = "red")+
stat_function(fun = function(x) .21, colour = "black")+
geom_point(x=0.3, y = .21)+
theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") +ylim(0,0.35)
```
---
class: center, middle
```{r, echo=FALSE, cache = TRUE, out.width = "80%"}
df <- data.frame(x = seq(0, 1, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) x-x^2, colour = "red")+
stat_function(fun = function(x) 0.1*x+0.18, colour = "black")+
geom_point(x=0.3, y = .21)+
theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") +ylim(0,0.35)
```
---
class: center, middle
```{r, echo=FALSE, cache = TRUE, out.width = "80%"}
df <- data.frame(x = seq(0, 1, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) x-x^2, colour = "red")+
stat_function(fun = function(x) 0.2*x+0.15, colour = "black")+
geom_point(x=0.3, y = .21)+
theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") +ylim(0,0.35)
```
---

class: center, middle
```{r, echo=FALSE,  warning=FALSE, cache = TRUE, out.width = "80%"}
df <- data.frame(x = seq(0, 1, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) x-x^2, colour = "red")+
stat_function(fun = function(x) 0.3*x+0.12, colour = "black")+
geom_point(x=0.3, y = .21)+
theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") +ylim(0,0.35)
```
---
class: center, middle
```{r, echo=FALSE, warning=FALSE, cache=TRUE, out.width = "80%"}
df <- data.frame(x = seq(0, 1, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) x-x^2, colour = "red")+
stat_function(fun = function(x) 0.4*x+.09, colour = "black")+
geom_point(x=0.3, y = .21)+
theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") +ylim(0,0.35)
```
---
## Derivatives as Limits

When the interval is wide, it is not a good approximation. But as the interval shrinks, the appoximation becomes better.

--

So, as you reduce the interval size to $0$, this line converges in the limit to the line that lies tangent to the curve at that point. Recall that $f(x) = x − x^2$ and consider a very small interval $\epsilon \dots$

--

\begin{align}
\frac{\Delta f(x)}{\Delta x} &= \lim_{\epsilon \to 0}\frac{f(x+\epsilon)-f(x)}{(x+\epsilon)-x}
\end{align}

---
## Derivatives as Limits


When the interval is wide, it is not a good approximation. But as the interval shrinks, the appoximation becomes better.

So, as you reduce the interval size to $0$, this line converges in the limit to the line that lies tangent to the curve at that point. 

Recall that $f(x) = x − x^2$ and consider a very small interval $\epsilon \dots$



\begin{align}
\frac{\Delta f(x)}{\Delta x} &= \lim_{\epsilon \to 0}\frac{f(x+\epsilon)-f(x)}{(x+\epsilon)-x} \\
&= \lim_{\epsilon \to 0} \frac{[(x+\epsilon)-(x+\epsilon)^2]-[(x)-(x)^2]}{(x+\epsilon)-x} 
\end{align}


---

## Derivatives as Limits


When the interval is wide, it is not a good approximation. But as the interval shrinks, the appoximation becomes better.

So, as you reduce the interval size to $0$, this line converges in the limit to the line that lies tangent to the curve at that point. 

Recall that $f(x) = x − x^2$ and consider a very small interval $\epsilon \dots$



\begin{align}
\frac{\Delta f(x)}{\Delta x} &= \lim_{\epsilon \to 0}\frac{f(x+\epsilon)-f(x)}{(x+\epsilon)-x} \\
&= \lim_{\epsilon \to 0} \frac{[(x+\epsilon)-(x+\epsilon)^2]-[(x)-(x)^2]}{(x+\epsilon)-x}  \\
&=\lim_{\epsilon \to 0} \frac{[x+\epsilon -(x^2+2\epsilon x+\epsilon^2)]-[(x)-(x)^2]}{(x+\epsilon)-x}  
\end{align}


---


## Derivatives as Limits


When the interval is wide, it is not a good approximation. But as the interval shrinks, the appoximation becomes better.

So, as you reduce the interval size to $0$, this line converges in the limit to the line that lies tangent to the curve at that point. 

Recall that $f(x) = x − x^2$ and consider a very small interval $\epsilon \dots$



\begin{align}
\frac{\Delta f(x)}{\Delta x} &= \lim_{\epsilon \to 0}\frac{f(x+\epsilon)-f(x)}{(x+\epsilon)-x} \\
&= \lim_{\epsilon \to 0} \frac{[(x+\epsilon)-(x+\epsilon)^2]-[(x)-(x)^2]}{(x+\epsilon)-x}  \\
&=\lim_{\epsilon \to 0} \frac{[x+\epsilon -(x^2+2\epsilon x+\epsilon^2)]-[(x)-(x)^2]}{(x+\epsilon)-x}   \\
&=\lim_{\epsilon \to 0} \frac{x+\epsilon -x^2-2\epsilon x-\epsilon^2 -x+x^2}{x+\epsilon-x}
\end{align}


---


## Derivatives as Limits


\begin{align}
\frac{\Delta f(x)}{\Delta x} &= \lim_{\epsilon \to 0} \frac{x+\epsilon -x^2-2\epsilon x-\epsilon^2 -x+x^2}{x+\epsilon-x}
\end{align}

---
## Derivatives as Limits


\begin{align}
\frac{\Delta f(x)}{\Delta x} &= \lim_{\epsilon \to 0} \frac{x+\epsilon -x^2-2\epsilon x-\epsilon^2 -x+x^2}{x+\epsilon-x} \\
&= \lim_{\epsilon \to 0} \frac{\epsilon- 2\epsilon x-\epsilon^2}{\epsilon}
\end{align}

---

## Derivatives as Limits


\begin{align}
\frac{\Delta f(x)}{\Delta x} &= \lim_{\epsilon \to 0} \frac{x+\epsilon -x^2-2\epsilon x-\epsilon^2 -x+x^2}{x+\epsilon-x} \\
&= \lim_{\epsilon \to 0} \frac{\epsilon- 2\epsilon x-\epsilon^2}{\epsilon} \\
 &= \lim_{\epsilon \to 0} 1- 2 x-\epsilon
\end{align}

---

## Derivatives as Limits


\begin{align}
\frac{\Delta f(x)}{\Delta x} &= \lim_{\epsilon \to 0} \frac{x+\epsilon -x^2-2\epsilon x-\epsilon^2 -x+x^2}{x+\epsilon-x} \\
&= \lim_{\epsilon \to 0} \frac{\epsilon- 2\epsilon x-\epsilon^2}{\epsilon} \\
&= \lim_{\epsilon \to 0} 1- 2 x-\epsilon \\
&= 1- 2x
\end{align}

---
## Derivative as Limits

.pull-left[
```{r, echo=FALSE, warning=FALSE, cache=TRUE}
df <- data.frame(x = seq(0, 1, by= 1))

ggplot(df, aes(x)) + 
stat_function(fun = function(x) x-x^2, colour = "red")+
stat_function(fun = function(x) 0.4*x+.09, colour = "black")+
geom_point(x=0.3, y = .21)+
theme_minimal()+
xlab("x") + ylab("y") +
geom_hline(yintercept = 0, colour = "black")+
geom_vline(xintercept = 0, colour = "black") +ylim(0,0.35)
```
]

.pull-right[
Using this formula, the slope of the curve at $x = .3$ (the point from the previous examples) is exactly:


\begin{align}
\text{slope} &= 1 - 2(.3) \\
&= 0.4
\end{align}


Or, if we want to find the point at which the slope is $0$ (rate of change is $0$):


\begin{align}
0 &= 1 - 2x \\
2x &= 1 \\
x &= 0.5
\end{align}
]

---
## Derivatives

Now, we can formally state that the derivative is equivalent to:

\begin{align}
\frac{d[f(x)]}{dx} = \lim_{\epsilon \to 0}\frac{f(x+\epsilon)-f(x)}{(x+\epsilon)-x}
\end{align}

--

Using this approach, we can find:

  - A general equation for the slope at any point
  - The exact value of the slope at a given point
  - The point that has a given slope
---

## Derivative notation

Different notation systems for expressing derivatives:

--

First derivative: $\frac{d[f(x)]}{dx}$, $\frac{d}{dx}[f(x)]$, $f'(x)$, or $f^1(x)$

--

Second derivative: $\frac{d^2[f(x)]}{dx^2}$, $\frac{d^2}{dx^2}[f(x)]$, $f''(x)$, or $f^2(x)$

--

We'll be using $f'(x)$ and $f''(x)$ but you are free to choose whatever makes sense to you!

---
## Cautionary Notes on Derivatives

A few assumptions in using this approach to find the slope: 

- The function is continuous (no gaps or jumps)
- The derivative exists (the limit of the slope is the same from the left as it is from the right), or *no sharp corners*.

--

For nearly all political science applications, these are fine assumptions. But it is important to state them explicitly and be aware that they’re there.

---

## Straightforward Derivatives

Fortunately (for humans), it is not necessary to take limits with $\epsilon$'s to find derivatives every time. A few rules generate the derivatives of most functions.

--

The first rule is the *power rule*.

--

If $f(x) = cx^n$, then its derivative is $f'(x) = ncx^{n-1}$.

--

The simplest example: consider the line $f(x) = 3x$. What's its derivative? 

--

\begin{align}
f'(x) &= (3x)' \\
      &= (3x^1)' \\
      &= 1 \times 3x^{1-1} \\
      &= 3x^0 = 3
\end{align}

--

A special case of the power rule is that the *derivative of a constant is zero*.

---
## Straightforward Derivatives

Let $f(x) = 3x^2$, then:

--
\begin{align}
f'(x) &= (2) 3x^{2-1} \\
      &= 6x \\
\end{align}

--
Let $g(x) = x^5$, then:

--
\begin{align}
g'(x) &= (5)x^{5-1} \\
      &= 5x^4 \\
\end{align}

--
Let $h(x) = 7x^{\frac{1}{2}}$, then:

--
\begin{align}
h'(x) &=\left(\frac{1}{2}\right)7x^{\frac{1}{2}-1} \\
      &= \frac{7}{2}x^{-\frac{1}{2}} \\
\end{align}

---
## Exercises

Find the following derivatives, and calculate the instantaneous slope of the curves at the point $x = 2$:

$f(x) = \frac{1}{4}x^4$

$g(x) = \frac{2}{x^3}$  [Hint: How else can we express fractions?]

$h(x) = 4x^{\frac{5}{2}}$

$j(x) = \sqrt[\leftroot{-2}\uproot{2}3]{x}$ [Hint: How else can we express roots?]

---
## Derivative of a Sum (or Difference)

The derivative of a sum (difference) is the sum (difference) of the derivatives.

--
\begin{align}
(f(x) +g(x))' = f'(x)+g'(x)
\end{align}

--
For example, let's consider $f(x) = 4x^2$ and $g(x) =5x^3$. What is $(f(x)+g(x))'$?

--
\begin{align}
f'(x)+g'(x) &= (4x^2)' + (5x^3)' \\
&= (2)4x^{2-1} + (3)5x^{3-1}\\
&= 8x +15x^2
\end{align}

--

Let $f(x) = x$ and $g(x) =x^2$.  What is $(f(x)-g(x))'$?

--

\begin{align}
f'(x)-g'(x) &= (x)' - (x^2)' \\
&= (1)x^{1-1} - (2)x^{2-1}\\
&= 1-2x
\end{align}

---
## Derivative of a Sum (or Difference)

Find the derivative of $h(x) = 5x^5-10x^3+6x^2-3$ and the rate of change when $x=1$.

--

\begin{align}
h'(x) &= (5x^5-10x^3+6x^2-3)' \\
      &= (5x^5)'-(10x^3)'+(6x^2)'-(3)' \\
      &= 5 \times 5x^{5-1} - 3 \times 10x^{3-1}+2\times 6x^{2-1} - 0 \\
      &= 25x^4 - 30x^2+12x
\end{align}

--

What is the rate of change when $x$ is equal to one?

--

\begin{align}
h'(1) &=  25x^4 - 30x^2+12x \\
h'(1) &=  25(1)^4 - 30(1)^2+12(1) \\
h'(1) &= 7
\end{align}
---

## Derivative of a Product

The derivative of a product two functions (let's say $f(x)$ and $g(x)$) is:

--

\begin{align}
(f(x) \times g(x))'= f'(x) \times g(x) + f(x) \times g'(x)
\end{align}

--

or the derivative of the first function *times* second function *plus* the first function *times* the derivative of second function.

--

This is **the product rule**.

---

## Derivative of a Product

As an example consider $f(x) =x^2 +1$ and $g(x) = x^3-4x$.  

What is $(f(x) \times g(x))'$?

--


\begin{align}
(f(x) \times g(x))' &= ((x^2 +1)(x^3-4x))' \\
&= (x^2 +1)'(x^3-4x)+ (x^2 +1)(x^3-4x)' \\
&= (2x)(x^3-4x)+ (x^2 +1)(3x^2-4) \\
&= 2x^4-8x^2+3x^4-4x^2+3x^2-4 \\
&= 5x^4-9x^2-4
\end{align}

--

This is easy to check by multiplying out the polynomial: $(x^2 +1)(x^3-4x) = x^5-3x^3-4x$, and then taking its derivative:

--
\begin{align}
(x^5-3x^3-4x)' = 5x^4-9x^2-4
\end{align}

---

## Derivative of a Quotient

The derivative of a quotient of two functions (let's say $f(x)$ and $g(x)$) is:

--
\begin{align}
\left(\frac{f(x)}{g(x)}\right)'= \frac{f'(x) \times g(x)-f(x) \times g'(x)}{g(x)^2}
\end{align}

--

or the bottom *times* the derivative of the top *minus* the top *times* the derivative of the bottom, all *divided* by the bottom *squared*.

--

This is referred to as **the quotient rule**.


---

## Derivative of a Quotient

Let $f(x) =x^2 +1$ and $g(x) = x^3-4x$. What is $\left(\frac{f(x)}{g(x)}\right)'$?

--


\begin{align}
\left(\frac{f(x)}{g(x)}\right)' &= \left(\frac{x^2 +1}{x^3-4x}\right)' \\
&= \frac{(x^2 +1)'(x^3-4x)-(x^2 +1)(x^3-4x)'}{(x^3-4x)^2} \\
&=  \frac{(2x)(x^3-4x)-(x^2 +1)(3x^2-4)}{(x^3-4x)^2} \\
&= \frac{2x^4-8x^2-(3x^4-4x^2+3x^2-4)}{(x^3-4x)^2}  \\
&= \frac{-x^4-7x^2+4}{(x^3-4x)^2}
\end{align}

---
## Exercises

Find the derivatives of the following expressions:

$(3x^2-4x+2)(x^3-x^2+x-1)$

$\frac{4x+1}{3x^2-2}$

---
## Derivatives of Nested Functions

Let's say $h(x) = f(g(x))$.

The derivative of one function nested inside another is:

--
\begin{align}
(f(g(x)))'= f'(g(x)) \times g'(x)
\end{align}

--

or the derivative of the outside with respect to the inside *times* the derivative of the inside function.
--

This is referred to as **the chain rule**.

--

This looks messy, but is actually fairly straightforward and extremely useful as a way to find derivatives of complex functions by treating them as nested chains of functions.


---

## Derivatives of Nested Functions

Let $h(x) = 6(3x^2 + 2)^4$. Observe that this can be thought of as two nested functions, such that $g(x) = 3x^2 + 2$ and $f(x) = 6x^4$, and $h(x) = f(g(x))$. 

What is $h'(x)$?

--

\begin{align}
h(x)' = (f(g(x)))' &= (6(3x^2 + 2)^4)' \\
&= (4)6(3x^2 + 2)^{4-1}(3x^2 + 2)' \\
&= 24(3x^2 + 2)^{3}(6x) \\
&= 144x(3x^2 + 2)^{3} \\
\end{align}

---

## Derivatives of Nested Functions

Let $k(x) = 3(6x^4)^2 + 2$. Observe that this can be thought of the same two functions nested in the reverse order, such that $g(x) = 3x^2 + 2$ and $f(x) = 6x^4$, and $k(x) = g(f(x))$.

What is $k'(x)$?

--

\begin{align}
k(x)' = (g(f(x)))' &= (3(6x^4)^2 + 2)' \\
&= (3(6x^4)^2)' + (2)' \\
&= (2) 3(6x^4)^{2-1}(6x^4)' + 0 \\
&=(2) 3(6x^4)^{2-1}(24x^{4-1}) \\
&=(2) 3(6x^4)(24x^{3}) \\
&= 864x^7
\end{align}

---

## Exercises

Express the functions below as the nested result of two simpler functions, and use the chain rule to find the derivative:

$(3x-1)^4$

$2(x^4+x^3)+7$

---
## Derivatives of Logarithms

The derivative for any logarithm base $b$ is 

--

\begin{align}
(\log_b(x))' = \frac{1}{ln(b)x}
\end{align}

--

It is important to note that a very special case of this is the derivative of a natural logarithm (or log base $e$), which is

--

\begin{align}
(log_e(x))' = (ln(x))' &= \frac{1}{ln(e)x} \\
&= \frac{1}{x}
\end{align}

---

## Derivatives of Logarithms

Let $f(x) = log_{10}(x)$. What is $f'(x)$?

--

\begin{align}
f'(x) = (log_{10}(x))' = \frac{1}{ln(10)x}
\end{align}

--

Let $g(x) = ln(3x^2+4)$. What is $g'(x)$? 

--

(using the chain rule):

\begin{align}
g'(x) = (ln(3x^2+4))' &= \frac{1}{3x^2+4}\times(3x^2+4)' \\
&= \frac{6x}{3x^2+4}
\end{align}
---

## Derivatives of Exponentials

The derivative for any exponential base $b$ is 

--

\begin{align}
(b^x)' = ln(b)b^x
\end{align}

--

It is important to note that a very special case of this is the derivative of $e^x$, which is

--

\begin{align}
(e^x)' &= ln(e)e^x \\
 &= 1 \times e^x \\
 &= e^x
\end{align}


---

## Derivatives of Exponentials

Let $f(x) = 4^x$. What is $f'(x)$?

--

\begin{align}
f'(x) =(4^x)' = ln(4)4^x
\end{align}

--

Let $g(x) = 2^{3x}$. What is $g'(x)$?

--

\begin{align}
g'(x) = (2^{3x})' &= ln(2)\times 2^{3x}\times(3x)' \\
&= 3ln(2)\times 2^{3x}
\end{align}

--

Let $h(x) = 4e^x$. What is $h'(x)$?

--

\begin{align}
h'(x) = (4e^x)' &= 4e^x 
\end{align}

---

## L'Hôpital's Rule

Sometimes when we calculate limits, it helps to take the derivative first.

--

Consider the limit: $\lim_{x \to 2} \frac{x^2 + x - 6}{x^2 - 4}$

$\frac{2^2 + 2 - 6}{2^2 - 4} = \frac{0}{0}$

If both the numerator and the denominator are $0$, $\infty$, or $-\infty$, we cannot evaluate the limit.

--

However, **L'Hôpital's Rule** says: $\lim_{x \to a} \frac{f(x)}{g(x)} = \lim_{x \to a} \frac{f'(x)}{g'(x)}$

Then, 

\begin{align}
\lim_{x \to 2} \frac{x^2 + x - 6}{x^2 - 4} 
&= \lim_{x \to 2} \frac{(x^2 + x - 6)'}{(x^2 - 4)'} \\
&= \lim_{x \to 2} \frac{2x + 1}{2x} \\
&= 1.25
\end{align}


---

class: center, inverse, middle


# End Day 3


```{css, echo=FALSE}
@media print {
  .has-continuation {
    display: block;
  }
}
```